---
title: "Food"
author: "Sharmarke, Evan & Mars"
date: "2023-2-24"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
pacman::p_load(tidyverse,readxl,shiny,dplyr,gapminder)

debugMode <- TRUE
showcodeMode <- FALSE

```

```{r fields processing, include=showcodeMode, message = debugMode, warning = debugMode}
CPIHistoricalData <- read_excel("historicalcpi.xlsx", skip = 1)
CPIHistoricalDataRaw <- CPIHistoricalData %>%
  remove_missing() %>%
  mutate(`Consumer Price Index item` = as.factor(`Consumer Price Index item`)) %>%
  mutate_if(is.character, as.numeric) %>% 
  rename(cpiItem = `Consumer Price Index item`)
#head(CPIHistoricalDataRaw)
CPIHistoricalDataRaw$cpiItem
CPIHistoricalDataRaw <- CPIHistoricalDataRaw %>% 
  mutate(level = case_when((cpiItem %in% c('All food'))~ 1,
                           (cpiItem %in% c('Food away from home','Food at home')) ~ 2,
                           (cpiItem %in% c('Meats, poultry, and fish', 'Eggs', 'Dairy products', 'Fats and oils', 'Fruits and vegetables', 'Sugar and sweets','Cereals and bakery products', 'Nonalcoholic beverages', 'Other foods')) ~ 3,
                           (cpiItem %in% c('Meats', 'Poultry', 'Fish and seafood', 'Fresh fruits and vegetables', 'Processed fruits and vegetables')) ~ 4,
                           (cpiItem %in% c('Beef and veal', 'Pork', 'Other meats', 'Fresh fruits', 'Fresh vegetables')) ~ 5,
                           TRUE ~ 0
                           ),
         level = as.factor(level)
         )
CPIHistoricalDataRaw %>% 
  select(cpiItem, level) %>%
  view()
#head(CPIHistoricalDataRaw)
CPIHistoricalLong <- CPIHistoricalDataRaw %>% 
  pivot_longer(cols = !c(cpiItem, level)) %>%
  rename(year = name, percentchange = value) %>% 
  mutate(percentchange = round(percentchange*0.01, 3))
#head(CPIHistoricalDataRaw)
CPIHistoricalDataFrame <- read_excel("historicalcpi.xlsx", sheet = "df")
head(CPIHistoricalDataFrame)
```


## Introduction

The consumer price index (CPI), which monitors the change in prices consumers pay for food, is very important for the nourishment and survival of our species. Our data was gathered by the USDA and showcases the CPI changes in food categories from 1974-2021. Understanding the changes in the CPI helps us look at food prices, insecurity, nutrition, and more. From our graphs, we would like to show the fluctuation in these food prices within each year, spanning from 1974-2021. As far as limitations, we assume that food is a global resource that is useful to everyone. However, we do not see that bias as a destructive force that changed our dashboard in any way. For methods and further information about the data, our group urges you to look at their website for more information found in the README file. 

## CPI Data {.tabset}

### Historical CPI

```{r historical, echo=showcodeMode, message = debugMode, warning = debugMode}
renderPlot({
  CPIHistoricalDataFrame %>%
    filter(item == "All food") %>%
  ggplot(mapping = aes(x = year,
                       y = percentChange,
                       fill = percentChange)) + 
  geom_col() + theme(aspect.ratio = 0.11)
})
  inputPanel(
  selectInput("cpiHistoricalItem", 
              label = "Item of showing in Graph",
              choices = unique(CPIHistoricalDataFrame$item),
              selected = "All food",
              multiple = TRUE),
  sliderInput("cpiHistoricalYear", label = "Year Range",
              min = 1974, max = 2021, value = c(1974,2021), step = 1, sep = ""),
)



renderPlot({
    CPIHistoricalDataFrame %>%
      filter(item %in% input$cpiHistoricalItem,
             year >= min(input$cpiHistoricalYear[1]), 
             year <= max(input$cpiHistoricalYear[2])) %>%
    ggplot(mapping = aes(x = year,
                         y = percentChange,
                         color = item)) + 
    geom_line() + guides(fill = TRUE) + 
    theme(legend.position = "left") + 
    labs(
      x = NULL,
      y = "Changes in Year",
      title = "Historical CPI"
  )
})
head(CPIHistoricalDataRaw)

```

### Forecast CPI

``` 
CPIForecastWide <- CPIForecastData %>% 
  remove_missing() %>% # remove the unused columns
  mutate(`Consumer Price Index item` = as.factor(`Consumer Price Index item`)) %>% #Convert the CPI items into factors
  mutate_if(is.character, as.numeric) %>% # Convert any columns that were assumed to be characters into numeric. We Could have specified the data types in the read_excel but this conversion is intuitive compared to multiple complex arguments to load data.
  rename(cpiItem = `Consumer Price Index item`) %>%#Rename the item column to something easier to code
  select(cpiItem, `Annual 2020`, `Annual 2021`, `Annual 2022`, ...10, )
  
colnames(CPIForecastWide) <- c('cpiItem','2020', '2021', '2022', '2023')
  

warnings() # view the warnings which show the blank rows being removed and the "N/A" values being converted into NA.

CPIForecastWide$cpiItem

CPIForecastWide <- CPIForecastWide %>%
  mutate(level = case_when((cpiItem %in% c('All food'))~ 1,
                           (cpiItem %in% c('Food away from home','Food at home')) ~ 2,
                           (cpiItem %in% c('Meats, poultry, and fish', 'Eggs', 'Dairy products', 'Fats and oils', 'Fruits and vegetables', 'Sugar and sweets','Cereals and bakery products', 'Nonalcoholic beverages', 'Other foods')) ~ 3,
                           (cpiItem %in% c('Meats', 'Poultry', 'Fish and seafood', 'Fresh fruits and vegetables', 'Processed fruits and vegetables')) ~ 4,
                           (cpiItem %in% c('Beef and veal', 'Pork', 'Other meats', 'Fresh fruits', 'Fresh vegetables')) ~ 5,
                           TRUE ~ 0
                           ),
         level = as.factor(level) #convert the new values into a factor
         )

CPIForecastWide %>% # check to make sure each of the categories is at their appropriate level
  select(cpiItem, level) %>%
  view()

CPIForecastLong <- CPIForecastWide %>% #Pivot the wide data set to minimize the number of columns, making it easier to visualize and create drill down functions.
  pivot_longer(cols = !c(cpiItem, level)) %>%
  rename(year = name, percentchange = value) %>% #rename the columns to meaningful variables.
  mutate(percentchange = round(percentchange*0.01, 3)) %>%# convert the percentage to an number that fits a percentage scale 0% to 100 % (0 to 1)
  filter(year %in% c('2020', '2021', '2022', '2023'))
```

```{r forecast, echo=showcodeMode, message = debugMode, warning = debugMode}
renderPlot({
  CPIForecastLong %>%
    filter(item == "All food") %>%
  ggplot(mapping = aes(x = year,
                       y = percentChange,
                       fill = percentChange)) + 
  geom_col() + theme(aspect.ratio = 0.11)
})
  inputPanel(
  selectInput("cpiForecastItem", 
              label = "Item of showing in Graph",
              choices = unique(CPIForecastLong$item),
              selected = "All food",
              multiple = TRUE))
  sliderInput("cpiForecastYear", label = "Year Range",
              min = 1974, max = 2021, value = c(1974,2021), step = 1, sep = "")
```
