---
title: "Food"
author: "Sharmarke, Evan & Mars"
date: "2023-2-24"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
pacman::p_load(tidyverse,readxl,shiny,dplyr,gapminder)

debugMode <- TRUE
showcodeMode <- FALSE

```

```{r fields processing, include=showcodeMode, message = debugMode, warning = debugMode}
#Historical Part
CPIHistoricalDataRaw <- read_excel("CPIHistorical.xlsx", sheet = "raw")
CPIHistoricalDataFrame <- read_excel("CPIHistorical.xlsx", sheet = "df")

#Forecast Part
CPIForecastDataRaw <- read_excel("CPIForecast.xlsx", sheet = "raw")
CPIForecastDataFrame <- read_excel("CPIForecast.xlsx", sheet = "df")

```


## Introduction

The consumer price index (CPI), which monitors the change in prices consumers pay for food, is very important for the nourishment and survival of our species. Our data was gathered by the USDA and showcases the CPI changes in food categories from 1974-2021. Understanding the changes in the CPI helps us look at food prices, insecurity, nutrition, and more. From our graphs, we would like to show the fluctuation in these food prices within each year, spanning from 1974-2021. As far as limitations, we assume that food is a global resource that is useful to everyone. However, we do not see that bias as a destructive force that changed our dashboard in any way. For methods and further information about the data, our group urges you to look at their website for more information found in the README file. 

In addition to our previous data, we included the forecast CPI data for these same food categories. We would like to use this to visualize future predictions for changes in CPI. 

## Consumer Price Index Data {.tabset}

### Historical CPI

```{r historical, echo=showcodeMode, message = debugMode, warning = debugMode}
renderPlot({
  CPIHistoricalDataFrame %>%
    filter(item == "All food") %>%
  ggplot(mapping = aes(x = year,
                       y = percentChange,
                       fill = percentChange)) + theme_minimal() + 
  geom_col() + theme(legend.position = "none")
})

inputPanel(
  selectInput("cpiHistoricalItem", 
              label = "Item of showing in Graph",
              #choices = unique(CPIForecastDataFrame$item),
              choices = list('All Foods' = list("All food"),
                             'Food in/out of Home' = list("Food away from home", "Food at home", "Eggs", "Dairy products", "Fats and oils"),
                             'Meats Poultry & Fish' = list("Meats, poultry, and fish", "Meats", "Beef and veal", "Pork", "Other meats"),
                             'Fruits & Vegetables' = list("Fruits and vegetables", "Fresh fruits and vegetables", "Fresh fruits", "Fresh vegetables", "Processed fruits and vegetables"),
                             'Miscellaneous' = list("Sugar and sweets", "Cereals and bakery products","Nonalcoholic beverages","Other foods")),
              selected = list("Food away from home", "Food at home"),
              multiple = TRUE),
  sliderInput("cpiHistoricalYear", label = "Year Range",
              min = 1974, max = 2021, value = c(1974,2021), step = 1, sep = ""),
)

renderPlot({
 CPIHistoricalDataFrame %>%
  filter(item %in% input$cpiHistoricalItem,
         year >= min(input$cpiHistoricalYear[1]), 
         year <= max(input$cpiHistoricalYear[2])) %>%
  ggplot(mapping = aes(x = year,
                       y = percentChange,
                       color = item)) + 
  geom_line() + geom_point() + guides(fill = TRUE) + 
  theme_minimal() + theme(legend.position = "left") +

  labs(x = NULL,
       y = "Changes in Year",
       title = "Historical CPI"
  )
})

renderDataTable({
  CPIHistoricalDataFrame %>%
    select(item, year, percentChange)
    
})

```

### Forecast CPI

```{r forecast, echo=showcodeMode, message = debugMode, warning = debugMode}
#renderPlot({
#  CPIForecastDataFrame %>%
#    filter(item == "All food") %>%
#  ggplot(mapping = aes(x = year,
#                       y = percentChange,
#                       fill = percentChange)) + theme_minimal() + 
#  geom_col() + theme(aspect.ratio = 0.11, legend.position = "none")
#})

inputPanel(
  selectInput("cpiForecastItem", 
              label = "Item of showing in Graph",
              #choices = unique(CPIForecastDataFrame$item),
              choices = list('All Foods' = list("All food"),
                             'Food in/out of Home' = list("Food away from home", "Food at home", "Eggs", "Dairy products", "Fats and oils"),
                             'Meats Poultry & Fish' = list("Meats, poultry, and fish", "Meats", "Beef and veal", "Pork", "Other meats"),
                             'Fruits & Vegetables' = list("Fruits and vegetables", "Fresh fruits and vegetables", "Fresh fruits", "Fresh vegetables", "Processed fruits and vegetables"),
                             'Miscellaneous' = list("Sugar and sweets", "Cereals and bakery products","Nonalcoholic beverages","Other foods")),
              selected = list("Food away from home", "Food at home"),
              multiple = TRUE)
)

renderPlot({
 CPIForecastDataFrame %>%
  filter(item %in% input$cpiForecastItem) %>%
  ggplot(mapping = aes(x = year,
                       y = percentChange,
                       fill = item)) + 
  geom_boxplot() + #guides(fill = TRUE) + 
  theme_minimal() + theme(legend.position = "left") +

  labs(x = NULL,
       y = "Changes in Year",
       title = "Forecast CPI"
  )
})

renderDataTable({
  CPIForecastDataFrame %>%
    select(item, year, percentChange)
    
})
```
