---
title: "forecasting-food-team-galactic"
author: "Sharmarke, Evan & Mars"
date: "2023-2-24"
output: html_document
runtime: shiny
---

```{r load data, include=FALSE}
# Sample Code to import data
library(readxl)
pacman::p_load(tidyverse,readxl,shiny)
CPIForecastData <- read_excel("CPIForecast.xlsx",
                              skip = 1)
CPIHistoricalData <- read_excel("historicalcpi.xlsx", 
                            skip = 1)
CPIHistoricalWide <- CPIHistoricalData %>% #Use the CPI historical data for piping the following functions and once complete, store this as new data object that is a wide format.
  remove_missing() %>% # remove the unused columns
  mutate(`Consumer Price Index item` = as.factor(`Consumer Price Index item`)) %>% #Convert the CPI items into factors
  mutate_if(is.character, as.numeric) %>% # Convert any columns that were assumed to be characters into numeric. We Could have specified the data types in the read_excel but this conversion is intuitive compared to multiple complex arguments to load data.
  rename(cpiItem = `Consumer Price Index item`) #Rename the item column to something easier to code
warnings() # view the warnings which show the blank rows being removed and the "N/A" values being converted into NA.
CPIHistoricalWide$cpiItem
CPIHistoricalWide <- CPIHistoricalWide %>% # Update the wide object using the hierarchical levels and named relationships.
  mutate(level = case_when((cpiItem %in% c('All food'))~ 1,
                           (cpiItem %in% c('Food away from home','Food at home')) ~ 2,
                           (cpiItem %in% c('Meats, poultry, and fish', 'Eggs', 'Dairy products', 'Fats and oils', 'Fruits and vegetables', 'Sugar and sweets','Cereals and bakery products', 'Nonalcoholic beverages', 'Other foods')) ~ 3,
                           (cpiItem %in% c('Meats', 'Poultry', 'Fish and seafood', 'Fresh fruits and vegetables', 'Processed fruits and vegetables')) ~ 4,
                           (cpiItem %in% c('Beef and veal', 'Pork', 'Other meats', 'Fresh fruits', 'Fresh vegetables')) ~ 5,
                           TRUE ~ 0
                           ),
         level = as.factor(level) #convert the new values into a factor
         )
CPIHistoricalWide %>% # check to make sure each of the categories is at their appropriate level
  select(cpiItem, level) %>%
  view()
CPIHistoricalLong <- CPIHistoricalWide %>% #Pivot the wide data set to minimize the number of columns, making it easier to visualize and create drill down functions.
  pivot_longer(cols = !c(cpiItem, level)) %>%
  rename(year = name, percentchange = value) %>% #rename the columns to meaningful variables.
  mutate(percentchange = round(percentchange*0.01, 3)) # convert the percentage to an number that fits a percentage scale 0% to 100 % (0 to 1)

```


```{r eruptions, echo=FALSE}
head(CPIHistoricalWide)
inputPanel(
  selectInput("n_breaks", label = "Number of bins:",
              choices = c(10, 20, 35, 50), selected = 20),
  
  sliderInput("bw_adjust", label = "Bandwidth adjustment:",
              min = 0.2, max = 2, value = 1, step = 0.2)
)
renderPlot({
  hist(faithful$eruptions, probability = TRUE, breaks = as.numeric(input$n_breaks),
       xlab = "Duration (minutes)", main = "Geyser eruption duration")
  
  dens <- density(faithful$eruptions, adjust = input$bw_adjust)
  lines(dens, col = "blue")
})
```

## Embedded ApplicatiThis R Markdown document is made interactive using Shiny. Unlike the more traditional workflow of creating static reports, you can now create documents that allow your readers to change the assumptions underlying your analysis and see the results immediately. 

To learn more, see [Inon

It's also possible to embed an entire Shiny application within an R Markdown document using the `shinyAppDir` function. This example embeds a Shiny application located in another directory:

```{r tabsets, echo=FALSE}
shinyAppDir(
  system.file("examples/06_tabsets", package = "shiny"),
  options = list(
    width = "100%", height = 550
  )
)
```

Note the use of the `height` parameter to determine how much vertical space the embedded application should occupy.

You can also use the `shinyApp` function to define an application inline rather then in an external directory.

In all of R code chunks above the `echo = FALSE` attribute is used. This is to prevent the R code within the chunk from rendering in the document alongside the Shiny components.
